Variable Selection in Convex Function Estimation

### 2013-11-12 tex/scam/additive_faithful.tex
more update

We say that $f$ \textbf{depends on} coordinate $k$ if, for all $x_k \in [0,1]$, the set 
$\big\{ x'_k \in [0, 1] \,:\, f(x_k, \mathbf{x}_{-k}) = f(x'_k, \mathbf{x}_{-k}) 
\trm{ for almost all  $\mathbf{x}_{-k}$} \big\}$ 
has measure strictly less than 1.\\

@@

@@

In the case of Theorem~\ref{thm:convex_faithful}, we can give some intuition: Lemma~\ref{lem:general_int_reduction}, we know that the
additive approximation zeroes out $k$ when, fixing $x_k$, every
``slice'' of $f$ integrates to zero. We prove
Theorem~\ref{thm:convex_faithful} by showing that ``slices'' of convex
functions that integrate to zero cannot be ``glued'' together while
still maintaining convexity.

We give the proof of Theorem~\ref{thm:convex_recall}, which also proves one direction of Theorem~\ref{thm:convex_faithful}. The other direction of Theorem~\ref{thm:convex_faithful} is trivial to show.

@@

By assumption, we know that there exist at least two disjoint closed intervals $R_0, R_1$ where $p(\mathbf{x}_{-k} \given x_k)$ is independent on $x_k$. That is, there exist density functions $\phi_0(\mathbf{x}_{-k})$ and $\phi_1(\mathbf{x}_{-k})$ such that $p(\mathbf{x}_{-k} \given x_k) = \phi_0(\mathbf{x}_{-k})$ for all $x_k \in R_0$ and $p(\mathbf{x}_{-k} \given x_k) = \phi_1(\mathbf{x}_{-k})$ for all $x_k \in R_1$.

@@

Since $x^0_k$ is in the interior of $R_0$, we use the bounded convergence theorem and conclude that $\int_{\mathbf{x}_{-k}} \phi_0( \mathbf{x}_{-k}) g(\mathbf{x}_{-k}) = 0$ as well.

@@

Now, for all $x_k \in R_0$, 
\[
\int_{\mathbf{x}_{-k}} \phi_0(\mathbf{x}_{-k}) 
     \Big(f(x_k, \mathbf{x}_{-k}) - f(x^0_k, \mathbf{x}_{-k}) 
     - g(\mathbf{x}_{-k}) (x_k - x^0_k) \Big) d\mathbf{x}_{-k} = 0
\]
Since $\phi_0$ is positive by assumption, we conclude that for all $x_k \in R_0$, for all $\mathbf{x}_{-k}$, 
$$
f(x_k, \mathbf{x}_{-k}) = f(x^0_k, \mathbf{x}_{-k}) - g(\mathbf{x}_{-k}) (x_k - x^0_k)
$$

>>>>

We say that $f$ \textbf{depends on} coordinate $k$ if, for all $x_k \in [0,1]$, the set 
$\big\{ x'_k \in [0, 1] \,:\, f(x_k, \mathbf{x}_{-k}) = f(x'_k, \mathbf{x}_{-k}) 
\trm{ for almost all  $\mathbf{x}_{-k}$} \big\}$ 
has probability strictly less than 1.\\

@@

\subsection{Additive Faithfulness of Convex Functions}

@@

We pause to give some intuition before we present the full proof: 
suppose the underlying distribution is a product distribution for a second, 
then we know from lemma~\ref{lem:general_int_reduction} that the
additive approximation zeroes out $k$ when, fixing $x_k$, every
``slice'' of $f$ integrates to zero. We prove
Theorem~\ref{thm:convex_faithful} by showing that ``slices'' of convex
functions that integrate to zero cannot be ``glued'' together while
still maintaining convexity.

We give the proof of Theorem~\ref{thm:convex_recall}, which also proves one direction of Theorem~\ref{thm:convex_faithful}. The other direction of Theorem~\ref{thm:convex_faithful} is trivial.

@@

Let us assume that for all $x_k$, $ \E[ f(X) - \sum_{k'} f_{k'}(X_{k'}) \given x_k] - \E f(X) = 0 $.\\

By the intermittent independence assumption, we know that there exists at least two disjoint closed intervals $R_0, R_1$ where $p(\mathbf{x}_{-k} \given x_k)$ is independent of $x_k$. That is, there exist density functions $\phi_0(\mathbf{x}_{-k})$ and $\phi_1(\mathbf{x}_{-k})$ such that $p(\mathbf{x}_{-k} \given x_k) = \phi_0(\mathbf{x}_{-k})$ for all $x_k \in R_0$ and $p(\mathbf{x}_{-k} \given x_k) = \phi_1(\mathbf{x}_{-k})$ for all $x_k \in R_1$.

@@

The last equality follows because we assumed that $ \E[ f(X) - \sum_{k'} f_{k'}(X_{k'}) \given x_k] - \E f(X) = 0 $ for all $x_k$. \\

Since $x^0_k$ is in the interior of $R_0$, and since the domain of $f$ is a compact set, we use the bounded convergence theorem and conclude that $\int_{\mathbf{x}_{-k}} \phi_0( \mathbf{x}_{-k}) g(\mathbf{x}_{-k}) = 0$ as well.\\

@@

Now, the critical observation we make is that, for all $x_k \in R_0$, 
\[
\int_{\mathbf{x}_{-k}} \phi_0(\mathbf{x}_{-k}) 
     \Big(f(x_k, \mathbf{x}_{-k}) - f(x^0_k, \mathbf{x}_{-k}) 
     - g(\mathbf{x}_{-k}) (x_k - x^0_k) \Big) d\mathbf{x}_{-k} = 0
\]
Since $\phi_0$ is positive by assumption and the second term is non-negative, we conclude that the second term must be exactly zero, that is, for all $x_k \in R_0$, for all $\mathbf{x}_{-k}$, 
$$
f(x_k, \mathbf{x}_{-k}) = f(x^0_k, \mathbf{x}_{-k}) - g(\mathbf{x}_{-k}) (x_k - x^0_k)
$$

### 2013-11-12 tex/scam/appendix.tex
more update

We will argue that all the components of the additive function $f^{*(s-1)}$ are also in $f^{*s}$. Let us denote the components of $f^{*s} = \sum_{k=1}^s f_k^*$. We will now invoke Lemma~\ref{lem:additive_component}, which is valid because we assume $X_1,...,X_s$ are independent by assumption A1'. By Lemma~\ref{lem:additive_component}, if we set $f^*_k = 0$, the resulting additive function $\sum_{k'\neq k} f_{k'}^*$ minimizes the population risk subject to the constraint that $f_k = 0$. By definition, $f^{*s}$ is $\argmin_k \sum_{k'\neq k} f_{k'}^*$ and thus share components with $f^{*s}$.

@@

Where we used the fact that $\E f^*_{k'} (X_{k'}) = 0$ for all $k'$ and the fact that $\E[ f_0(X) - \mu \,|\, X_k] = f^*_k(X_k)$ (Lemma~\ref{lem:int_reduction}).

>>>>

We will argue that all the components of the additive function $f^{*(s-1)}$ are also in $f^{*s}$. Let us denote the components of $f^{*s} = \sum_{k=1}^s f_k^*$. We will now invoke Corollary~\ref{cor:product_int_reduction}, which is valid because we assume $X_1,...,X_s$ are independent by assumption A1'. By Corollary~\ref{cor:product_int_reduction}, if we set $f^*_k = 0$, the resulting additive function $\sum_{k'\neq k} f_{k'}^*$ minimizes the population risk subject to the constraint that $f_k = 0$. By definition, $f^{*s}$ is $\argmin_k \sum_{k'\neq k} f_{k'}^*$ and thus share components with $f^{*s}$.

@@

Where we used the fact that $\E f^*_{k'} (X_{k'}) = 0$ for all $k'$ and the fact that $\E[ f_0(X) - \mu \,|\, X_k] = f^*_k(X_k)$ (Lemma~\ref{lem:general_int_reduction}).

### 2013-11-12 tex/scam/sparsistency.tex
more update

\begin{packed_enum}
\item Let $F$ be a distribution supported and positive on $\mathcal{X}=[-b,b]^p$. Let $X^{(1)},..., X^{(n)} \sim F$ be iid.
\item Let $Y = f_0(X) + \epsilon$ where $\epsilon$ is zero-mean noise. Let $Y^{(1)},...,Y^{(n)}$ be iid.
\item Let $S = \{1,...,s\}$ denote the relevant variables where $s\leq p$, i.e.,
  $f_0(X) = f_0(X_S)$.
\item Let $f_{01},...,f_{0s} \coloneqq \argmin_{f_1,...,f_s} \{ \E(f_0(X) - \sum_{k=1}^s f_k(X_k))^2 \,|\, \E[f_k(X_k)] = 0 \}$.
\end{packed_enum}

Each of our theorems will use a subset of the following assumptions:
\begin{packed_enum}
\item[A1:] $X_S, X_{S^c}$ are independent.  \ A1': $\{ X_k \}_{k \in S}$ are independent.
\item[A2:] $\|f_0\|_\infty \leq sB$ \  A2': $f_0$ is convex,
  twice-differentiable, $L$-Lipschitz, and $\trm{supp}(f_0) = S$.
\item[A3:] Suppose $\epsilon$ is mean-zero sub-Gaussian, independent of $X$, with sub-Gaussian scale $\sigma$.
\item[A4:] For all $k=1,...,s$, $\E(f_{0k}(X_k))^2 \geq \alpha$ for some positive constant $\alpha$.
\end{packed_enum}

We will use assumptions A1, A2, A3 to control the probability of false positives and the stronger assumptions A1', A2', A3, A4 to control the probability of false negatives. The $\alpha$ in assumption A4 can be interpreted as the signal strength of the relevant variable detection problem.

@@

@@

The proof of Theorem~\ref{thm:false_positive} exploits independence of
$\hat{r}$ and $\Delta_{k,j}$ from A1, and then uses concentration of
measure results to argue that $| \frac{1}{n} \hat{r}^\tran
\Delta_{k,j}|$ concentrates around zero at a desired rate. The fact
that $\hat{r}$ is a centered vector is crucial to our proof, and our
theory thus further illustrates the importance of imposing the
centering constraints in optimization \eqref{opt:alternate_opt}. Our
proof uses the concentration of the average of
data sampled \emph{without} replacement
\cite{serfling1974probability}, illustrating that the proof method is not a
trivial application of existing techniques. The full proof of
Theorem~\ref{thm:false_positive} is in
Section~\ref{sec:false_positive_proof} of the Appendix.

\begin{theorem} (Controlling false negatives)
\label{thm:false_negative}
Suppose assumptions A1', A2', A3 hold. Let $\ds \hat{f} = \{ \hat{d}_k, \hat{c}_k\}_{k\in S}$ be any solution to the restricted regression with both the $B$-boundedness and $L$-Lipschitz constraint. Let $c,C$ be absolute constants.
Suppose $L \max \left(\lambda_n, b (B+\sigma)B\sigma \sqrt{\frac{s^5}{n^{4/5}} \log sn} \right) \rightarrow 0$.
Then, for sufficiently large $n$, $\hat{f}_k = (\hat{d}_k, \hat{c}_k)
\neq 0$ for all $k \in S$ with probability at least $1-\frac{C}{n}$.
\end{theorem}

This is a finite sample version of
Theorem~\ref{thm:convex_faithful}. We need stronger assumptions in
Theorem~\ref{thm:false_negative} to use our additive faithfulness
result, Theorem~\ref{thm:convex_faithful}. We also include an extra
Lipschitz constraint so that we can use existing covering number
results \cite{Bronshtein:76}. Recent work
\cite{Guntu:13} shows that the Lipschitz constraint
is not required with more advanced empirical process theory
techniques. We give the full proof of Theorem~\ref{thm:false_negative}
in Section~\ref{sec:false_negative_proof} of the Appendix.

Combining Theorem~\ref{thm:false_positive} and
~\ref{thm:false_negative} and ignoring dependencies on $b,B,L,\sigma$,
we have the following result.
\begin{corollary}
  Assume A1', A2', A3. Let $\lambda_n = \Theta\left( \sqrt{
  \frac{s^3}{n} \log n \log (pn)} \right)$. Suppose $\lambda_n
  \rightarrow 0$ and $\sqrt{\frac{s^5}{n^{4/5}} \log sn} \rightarrow
  0$. Let $\hat{f_n}$ be a solution to (\ref{opt:alternate_opt}) with
  boundedness and Lipschitz constraints. Then 
  $\P( \trm{supp}(\hat{f_n}) = \trm{supp}(f_0) ) \rightarrow 1$.
\end{corollary}
The above corollary implies that sparsistency is achievable at the same exponential scaling of the ambient dimension $p = O(\exp(n^c)), c<1$ rate as parametric models. The cost of nonparametric modeling is reflected in the scaling with respect to $s$, which can only scale at $o(n^{4/25})$.

@@

>>>>

\begin{packed_enum}
\item Let $F$ be a distribution supported and positive on $\mathcal{X}=[-b,b]^p$. Let $X^{(1)},..., X^{(n)} \sim F$ be iid.
\item Let $Y = f_0(X) + \epsilon$ where $\epsilon$ is zero-mean noise. Let $Y^{(1)},...,Y^{(n)}$ be iid.
\item Let $S = \{1,...,s\}$ denote the relevant variables where $s\leq p$, i.e.,
  $f_0(X) = f_0(X_S)$.
\item Let $f^*_1,...,f^*_s \coloneqq \argmin_{f_1,...,f_s} \{ \E\Big(f_0(X) -\E f_0(X) - \sum_{k=1}^s f_k(X_k)\Big)^2 \,|\, \E[f_k(X_k)] = 0 \}$.
\end{packed_enum}

Each of our theorems will use a subset of the following assumptions:
\begin{packed_enum}
\item[A1:] $X_S, X_{S^c}$ are independent.  \ A1': $\{ X_k \}_{k \in S}$ are independent.
\item[A2:] $\|f_0\|_\infty \leq sB$ \  A2': $f_0$ is convex,
  twice-differentiable, and $L$-Lipschitz.
\item[A3:] Suppose $\epsilon$ is mean-zero sub-Gaussian, independent of $X$, with sub-Gaussian scale $\sigma$, i.e. for all $t \in \R$, $\E e^{t \epsilon} \leq e^{\sigma^2 t^2 / 2}$.
\item[A4:] For all $k=1,...,s$, $\E(f^*_s(X_k))^2 \geq \alpha$ for some positive constant $\alpha$.
\end{packed_enum}

We will use assumptions A1, A2, A3 to control the probability of false
positives and the stronger assumptions A1', A2', A3, A4 to control the
probability of false negatives.  Assumption A4 can be
weakened so that the relevant functions satisfy
$\E(f^*_s(X_k))^2 \geq \alpha_n$ for $\alpha_n$ decaying to zero 
at an appropriate rate.

@@

\begin{remark}
Assumption A4 ensures that the relevant variables are ``relevant enough''. Under A4, the population risk of an additive function with $s-1$ components is at least $\alpha$ larger than the population risk of the optimal additive function with $s$ components. See lemma~\ref{lem:minus_one_risk_increase} in section~\ref{sec:false_negative_proof} of the appendix.
\end{remark}

@@

The proof of Theorem~\ref{thm:false_positive} exploits independence of
$\hat{r}$ and $\Delta_{k,j}$ from A1, and then uses concentration of
measure results to argue that $| \frac{1}{n} \hat{r}^\tran
\Delta_{k,j}|$ concentrates around zero at a desired rate. The fact
that $\hat{r}$ is a centered vector is crucial to our proof, and our
theory thus further illustrates the importance of imposing the
centering constraints in optimization \eqref{opt:alternate_opt}. Our
proof uses the concentration of the average of
data sampled \emph{without} replacement
\cite{serfling1974probability}. The full proof of
Theorem~\ref{thm:false_positive} is in
Section~\ref{sec:false_positive_proof} of the Appendix.

\begin{theorem} (Controlling false negatives)
\label{thm:false_negative}
Suppose assumptions A1', A2', A3, A4 hold. Let $\ds \hat{f} = \{ \hat{d}_k, \hat{c}_k\}_{k\in S}$ be any solution to the restricted regression with both the $B$-boundedness and $L$-Lipschitz constraint. Let $c,C$ be absolute constants.
Suppose $s L \lambda_n \rightarrow 0$ and $Lb(sB+\sigma)sB \sqrt{\frac{s}{n^{4/5}} \log sn} \rightarrow 0$.
Then, for sufficiently large $n$, $\hat{f}_k = (\hat{d}_k, \hat{c}_k)
\neq 0$ for all $k \in S$ with probability at least $1-\frac{C}{n}$.
\end{theorem}

This is a finite sample version of
Theorem~\ref{thm:convex_faithful}. We need stronger assumptions in
Theorem~\ref{thm:false_negative} to use our additive faithfulness
result, Theorem~\ref{thm:convex_faithful}. We also include an extra
Lipschitz constraint so that we can use existing covering number
results \cite{Bronshtein:76}. Recent work
\cite{Guntu:13} shows that the Lipschitz constraint
is not required with more advanced empirical process theory
techniques; we leave the incorporation of this development as future work. 
We give the full proof of Theorem~\ref{thm:false_negative}
in Section~\ref{sec:false_negative_proof} of the Appendix.

Combining Theorem~\ref{thm:false_positive} and
~\ref{thm:false_negative} and ignoring dependencies on $b,B,L,\sigma$,
we have the following result.
\begin{corollary}
  Assume A1', A2', A3, A4. Let $\lambda_n = \Theta\left( \sqrt{
  \frac{s^3}{n} \log n \log (pn)} \right)$. Suppose $s \lambda_n
  \rightarrow 0$ and $\sqrt{\frac{s^5}{n^{4/5}} \log sn} \rightarrow
  0$. Let $\hat{f_n}$ be a solution to (\ref{opt:alternate_opt}) with
  boundedness and Lipschitz constraints. Then 
  $\P( \trm{supp}(\hat{f_n}) = \trm{supp}(f_0) ) \rightarrow 1$.
\end{corollary}
The above corollary implies that sparsistency is achievable at the same exponential scaling of the ambient dimension $p = O(\exp(n^c)), c<1$ rate as parametric models. The cost of nonparametric modeling is reflected in the scaling with respect to $s$, which can only scale at $o(n^{4/25})$.

\begin{remark}
\citet{dalalyan:12} have shown that under tradtional smoothness constraints, variable selection is achievable only if $n > O(e^s)$. It is interesting to observe that because of additive faithfulness, the convexity assumption enables a much better scaling of $n = O(\textrm{poly}(s))$, demonstrating that geometric constraints can be quite different from the previously studied smoothness conditions.
\end{remark}

@@

% DO NOT CHANGE; RefTex variables -minx

%%% Local Variables: ***
%%% mode:latex ***
%%% TeX-master: "paper.tex" ***
%%% End: ***
